---
title: "Using ggplot2 Inside data.table"
author: "John"
date: '2019-10-24'
slug: using-ggplot2-inside-data-table
tags:
- data.table
- ggplot2
- ggplot
- R
categories: []
---

```{r global_options, include=FALSE}
library(knitr)
knitr::opts_chunk$set(fig.pos = '!h')
```

I'm learning `data.table`. Recently I needed to plot a histogram of some particular columns. I already learned to use `.SDcols` and `.SD` to get at particular columns and subsets, but I couldn't figure out how to refer to the column's name in the plot itself.

Eventually I settled on the approach of mapping over the _indices_ of the data set and getting the correct column data and name during each `lapply` iteration.

Let me know your ideas for improvement. Like I said, I'm still learning `data.table`, so I'm sure I have much to improve. Here's my code:

```{r define_function}
histogram_my_columns <- function(data, columns = NULL) {
  require(data.table)
  require(ggplot2)
  if (!is.data.table(data)) {
    data <- as.data.table(data)
  }
  if (is.null(columns)) {
    columns <- names(data)
  }
  stopifnot(all(columns %in% names(data)))
  plotter <- function(index) {
    column <- data[[index]]
    column_name <- columns[index]
    p <- 
      ggplot() +
      geom_histogram(aes(column)) +
      ggtitle(paste0('Histogram of ', column_name))
  suppressMessages(print(p))
  TRUE
  }
  res <- data[, lapply(seq_along(.SD), plotter), .SDcols = columns]
}
```

```{r call_function}
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(ggplot2))
histogram_my_columns(mtcars, c('mpg', 'cyl', 'disp'))
```